---
title: "Reproducible Research: Peer Assessment 1"
author: "Cahit Bagdelen"
date: "July 19, 2015"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### 1. Load the data (i.e. read.csv())

```{r}
act <- read.csv(unz("activity.zip", "activity.csv"))
str(act)
```

### 2. Process/transform the data
Convert act$date to Date format:
```{r}
act$date <- as.Date(act$date)
str(act)
```
The summary of the activity data:
```{r}
summary(act)
```

## What is mean total number of steps taken per day?

### 1. Histogram of the total number of steps taken each day
```{r}
totalStepsPerDay <- aggregate(act$steps, by = list(act$date), sum, na.rm = TRUE )
totalStepsPerDay$Date <- as.Date(totalStepsPerDay$Group.1)
library("ggplot2")
ggplot(totalStepsPerDay, aes(x = totalStepsPerDay$Date, y = totalStepsPerDay$x)) + 
    geom_bar(stat="identity") + 
    xlab("Date") + 
    ylab("average number of steps per day") +
    ggtitle("Histogram of daily total number of steps")
```

### 2. Mean and median total number of steps taken per day
The mean total number of steps taken per day:
```{r}
meanStepsPerDay <- aggregate(act$steps, by = list(act$date), mean, na.rm = TRUE )
```

The median total number of steps taken per day:
```{r}
medianStepsPerDay <- aggregate(act$steps, by = list(act$date), median, na.rm = TRUE )
```

## What is the average daily activity pattern?

### 1. Make a time series plot 
time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
meanStepsPerInterval <- aggregate(act$steps, by = list(act$interval), mean, na.rm = TRUE )
str(meanStepsPerInterval)

ggplot(meanStepsPerInterval, aes(x = meanStepsPerInterval$Group.1, y = meanStepsPerInterval$x)) +
    geom_line() + 
    xlab("Intervals (5 min.)") + 
    ylab("mean daily total number of steps") +
    ggtitle("Timeseries for the mean daily total number of steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Maximum number of steps:
```{r}
max(meanStepsPerInterval$x)
```
The interval containing the maximum number of steps: 
```{r}
meanStepsPerInterval$Group.1[meanStepsPerInterval$x == max(meanStepsPerInterval$x)]
```

## Imputing missing values

### 1. total number of missing values in the dataset
the total number of rows with NAs:
```{r}
sum(!complete.cases(act))
```
the number of NAs in each column of the data frame:
```{r}
sum(is.na(act$date))
sum(is.na(act$steps))
sum(is.na(act$interval))
str(act)
```

### 2. Devise a strategy for filling in all of the missing values in the dataset.
Mean of the interval (across all days) can be used to fill the NAs for an interval at a specific date.
As shown above, interval means can be calculated as follows:
```{r}
meanStepsPerInterval <- aggregate(act$steps, by = list(act$interval), mean, na.rm = TRUE )
```

### 3. Fill missing values:
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
act_cmp <- act
n <- nrow(act)
for (i in 1:n) {
    if (is.na(act$steps[i])) {
        interval <- act$interval[i]
        mean <- meanStepsPerInterval$x[meanStepsPerInterval$Group.1 == interval]
        act_cmp$steps[i] <- mean
    }
}
```

Check number of NAs in the new data frame:
```{r}
sum(is.na(act_cmp$steps))
sum(!complete.cases(act_cmp))
str(act_cmp)
```

### 4. Histogram of the total number of steps taken each day
```{r}
totalStepsPerDay <- aggregate(act_cmp$steps, by = list(act_cmp$date), sum, na.rm = TRUE )
totalStepsPerDay$Date <- as.Date(totalStepsPerDay$Group.1)
library("ggplot2")
ggplot(totalStepsPerDay, aes(x = totalStepsPerDay$Date, y = totalStepsPerDay$x)) + 
    geom_bar(stat="identity") + 
    xlab("Date") + 
    ylab("average number of steps per day") +
    ggtitle("Histogram of daily total number of steps (after imputing NAs)")
```
As we can observe in the histogram, we have values for each interval, now.

## Are there differences in activity patterns between weekdays and week- ends?

### 1. Create a new factor variable in the dataset
Add weekDay as additional factor variable into the dataframe:
```{r}
weekDays <- weekdays(as.Date(act_cmp$date))
act_cmp$weekDay <- factor(weekDays != "Saturday" & weekDays != "Sunday", labels = c("weekend", "weekday"))
str(act_cmp)
``` 

### 2. Make a panel plot
panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
meanStepsPerInterval <- aggregate(act_cmp$steps, by = list(act_cmp$interval, act_cmp$weekDay), mean, na.rm = TRUE )
qplot(Group.1, x, data=meanStepsPerInterval, facets = Group.2 ~ ., geom = "line") + 
    xlab("Intervals (5 min)") + 
    ylab("mean total number of steps per day") + 
    ggtitle("Mean daily total number of steps on Weekdays & Weekends")

```
